Analysis-ready VCF at Biobank scale using Zarr

### 2024-05-10 paper.tex
A paragraph on Python data science
Closes #6
Closes #88

\section{Background}
Variant Call Format (VCF) is the standard format for interchanging genetic
variation data~\citep{danecek2011variant}, encoding information about 
DNA sequence polymorphisms among a set of samples with associated 
quality control metrics and metadata. 
Originally defined specifically as a text file,
it has been refined and standardised~\citep{rehm2021ga4gh} and the 
underlying data-model is now deeply embedded in bioinformatics practice.
Dataset sizes have grown explosively since the introduction of 
VCF as part of 1000 Genomes project~\citep{10002015global},
with Biobank scale initiatives such as 
Genomics England~\cite{turnbull2018100},
UK Biobank~\citep{bycroft2018genome,backman2021exome,halldorsson2022sequences,uk2023whole},
and the All of Us research program~\citep{all2024genomic}
collecting genome sequence data for hundreds of thousands of humans.
VCF's simple text-based design and widespread support makes it an 
excellent archival format, but it is an inefficient basis for analysis.
Methods that require efficient access to genotype data
either require conversion to the
PLINK~\cite{purcell2007plink,chang2015second} 
or BGEN~\citep{band2018bgen} 
formats~\citep[e.g.][]{yang2011gcta,mbatchou2021computationally,loh2015efficient}
or use 
bespoke binary formats~\citep[e.g.][]{
% Uses custom "bref3" format,
% https://faculty.washington.edu/browning/beagle/bref3.24May18.pdf
browning2018one, 
% .samples Zarr format
kelleher2019inferring,
% Has a "xcftools" package, but it still looks pretty experimental
hofmeister2023accurate} 
that support the required access patterns.
While PLINK and BGEN support efficient access to genotype data, neither
can accommodate the full flexibility of the VCF data model and conversion
is lossy.
PLINK's approach of storing the genotype matrix in uncompressed
packed-binary format provides efficient access to genotype 
data, but file sizes are substantially larger than 
the equivalent compressed VCF. 
For example, at two bits 
per genotype, the full genotype matrix for the GraphTyper SNP dataset 
in the 500K UKB WGS data~\citep{uk2023whole} is 116 TiB.
% 1,037,556,156 SNPs x  490,640 samples
% humanize.naturalsize(1_037_556_156 * 490_640 / 4, binary=True)
% '115.7 TiB'

Processing of Biobank scale datasets can be split into a 
few broad categories. The most basic analysis 
is quality control (QC). Variant QC is an 
involved and multi-faceted 
task [CITATIONS?], often requiring interactive, exploratory analysis
and incurring substantial computation over multiple QC fields.
Genotype calls are often refined via statistical methods,
% methods~\citep[e.g.][MORE and CHECK]{browning2021fast,browning2023statistical}
for example by phasing~[CITES] and imputation[CITES], 
creating additional copies of the dataset.
A common task to perform after QC is 
a genome wide association study (GWAS)~\cite{uffelmann2021genome}. 
The majority of tools for 
performing GWAS and related analyses require
data to be in PLINK format~\cite[e.g][]{chang2015second} [MORE],
and so data must be ``hard-called'' according to some QC criteria
and exported to additional copies.
Finally, variation datasets are often queried in exploratory
analyses, to find regions or samples of interest for a particular 
study (see section XXX for an example).

VCF cannot support any of these workflows efficiently
at the Biobank scale, and workarounds of copying subsets 
of data (i.e., discarding QC metrics in VCF) or copying to 
PLINK format have reached their limits.
The most intrinsically limiting aspect of VCF's design 
is its row-wise layout of data, which means that (for example)
information for a particular sample or field cannot be obtained without
retrieving the entire dataset.
The file-oriented paradigm is also unsuited to the realities 
of modern datasets, which are too large to download and 
often required to stay in-situ by data-access agreements.
Large files are stored in cloud environments, where the 
file systems that are required by classical file-oriented tools
are expensively emulated on the basic building blocks
of object storage. 
These multiple layers of inefficiences around processing
VCF data at scale in the cloud mean that it is 
time-consuming and expensive, and these vast datasets are 
not utilised to their full potential.

To achieve this full potential we 
need a new generation of tools that operate directly
on a primary data representation that supports 
efficient access across a range of applications,
with native support for cloud object storage.
Such a representation can be termed ``analysis-ready''
and ``cloud-native''~\citep{abernathey2021cloud}.
For the representation to be FAIR~\citep{wilkinson2016fair},
it must also be \emph{accessible}, using protocols that are 
``open, free, and universally implementable''.
There is no currently no efficient, FAIR representation of genetic variation
data suitable for cloud deployments.
Hail~\cite{ganna2016ultra,hail2024} has become the dominant platform
for quality control of large-scale variation dataset, 
and has been instrumental in the delivery of datasets such as 
gnomadAD~\cite{karczewski2020mutational,chen2024genomic}.
While Hail is built on open components
from the Hadoop distributed computing ecosytem~\citep{white2012hadoop},
it is complex, vertically integrated software,
and the custom storage formats are not intended for reuse.
Similarly, commerical solutions that have emerged to facilitate
the analysis of large-scale genetic variation data are either
based on proprietary~[Terra? BigQuery?]
or single-vendor technologies~\cite[e.g.][]{tiledb2024}.
The next generation of VCF analysis methods requires
an open, free and transparent data representation 
with multiple independent implementations.

In this article, we decouple the VCF data model from its row-oriented
file definition, and show how the data can be 
efficiently stored and analysed in a cloud-native, FAIR manner.
We do this by translating VCF data into Zarr format,
a method of storing large-scale multidimensional data as a regular
grid of compressed chunks. 
Zarr's elegant simplicity and first-class support for 
cloud object stores have led to 
it gaining substantial traction
across the sciences, and it is now used in multiple petabyte-scale
datasets in cloud deployments (see Methods for details).
We present the VCF Zarr specification which formalises this 
mapping, and the \texttt{vcf2zarr} 
utility which reliably converts large-scale VCFs into Zarr 
format. We show that VCF Zarr is much more compact than 
VCF and is competitive with state-of-the-art
file-based VCF compression tools. 
Moreover, we show that Zarr's storage of data in an analysis-ready 
format greatly facilitates computation,
with various benchmarks being substantially faster than
\texttt{bcftools} based pipelines, and again competetive
with state-of-the-art file-oriented methods. Finally, we show the 
utility of VCF Zarr on the Genomics England aggV2 dataset,
demonstrating that common \texttt{bcftools} queries can be performed orders
of magnitude more quickly using simple Python scripts.

@@

\section{Discussion}
The introduction of VCF as part of the work of the 1000 Genomes
project~\cite{10002015global} has transformed a fragmented 
landscape of mutually incompatible per-program file formats
into an ecosystem of tools that largely
interoperate~\cite{garrison2022spectrum}. In this 
sense VCF has been a huge success, and the underlying data model
is now deeply embedded in bioinformatics practice. As we have 
illustrated here, the shortcomings of the VCF and BCF file
formats (and more generally \emph{any} row-wise encoding of the 
data model) in terms of computational accessibility are severe,
and there can be little doubt that these file formats 
are incurring a large economic cost and 
holding back progress. 
As we have illustrated in the Genomics England
example here, these shortcomings are most acutely felt in 
large-scale data human datasets but any setting in which we 
wish to extract columns of the data from a row-wise 
VCF format will be inefficient 
when datasets are larger than a few hundred megabytes.

As we have demonstrated, the columnar approach, based on the 
popular Zarr standard and widely used compression libraries, 
can support many different types of data access
patterm efficiently, and has excellent computational accessibility.
In this approach there are some significant departures from 
standard practice, which warrant some discussion here.

Perhaps the largest deviation in the approach that we advocate here 
from standard practice is to decouple our concept of a dataset from
a file. Each VCF is broken up into a set of Zarr arrays, each 
corresponding to a large number of chunks stored in a directory
hierarchy (Zarr also supports other forms of chunk storage 
including single files and key-value stores however; see Methods).
While there are some disadvantages to large numbers of small files,
we argue that the benefits greatly outweigh them, particularly 
in the large, centrally controlled dataset setting. 
Firstly, this approach is inherently parallelisable, and chunks across the 
dataset written and read in parallel without needing complex 
synchronisation and prior layout planning. 
Such parallel write access is very difficult 
in the single-file setting, and one of the key weaknesses 
of the HDF5 format~\citep{folk2011overview} [CITATION with 
parallel troubles from HDF5?]
Secondly, mapping chunks in arrays to individually addressed
keys maps well to large-scale cloud object storage. Indeed,
Zarr is specifically designed for such usage, and 
several petabyte scale datasets have been successfully deployed
and used [citations?].
We have focused here on the classical single-server and 
HPC cluster settings, but a major advantage of Zarr is 
its cloud readiness.
% Probably getting into the weeds too much.
% Some large cloud-hosted VCF datasets are currently stored in 
% filesystems that emulate POSIX semantics on top of such object 
% stores, at a substantial cost in performance. Directly hosting 
% these datasets in object stores would have substantial 
% benefits, both in terms of storage costs and performance.
Thirdly, breaking up fields into different directories (or 
cloud buckets) enables fine-grained access control to the 
dataset. For example, all users may be allowed access to
variant-level data, but only a subset permitted to access
call-level individual data. Such access controls can be 
straightforwardly applied using file or cloud-hosting 
methods on the relevant directores/buckets. Such concerns
are entirely decoupled from the file format.

% This is poorly explained
Along with the move away from the single-file-storage paradigm,
the approach we outline here also suggests a shift from 
the classical Unix pipeline-based processing approach 
(but see below for a discussion of how existing pipelines can
be supported). Datasets flowing through pipelines inherently 
implies creating muliple copies of the data, and when 
multiple terabytes of either unrelated or unchanged data 
get copied several times through these pipes large inefficiencies 
are incurred (see Fig~\ref{fig-whole-matrix-compute}). 
We may think of this strategy as ``copy-oriented'' workflows.
A more efficient way to structure such multilayered computations
is in a  ``mask-oriented'' way. For example, to implement some
complex multilayered call filtering strategy we could imagine sequentially
updating a given per-call mask. Masks are cheap (e.g., the 
standard genotype missingness mask in~\ref{tab-genomics-england-data}
uses less than 1\% of the overall storage).

@@

Nonetheless, at least in the single-shared-copy model currently
used by most large-scale variation datasets, Zarr provides
an excellent basis for building a new generation of tools 
that compute \emph{directly} on the data, efficiently
accessing data by variant and by sample.
[Mention ``Cloud native data repositories''~\cite{abernathey2021cloud}]
The VCF Zarr specification and scalable vcf2zarr conversion utility
that we have provided here are the starting points of such 
an ecosystem, but much more would be needed to make a viable 
alternative for everyday bioinformatics workflows. However, a few key
pieces of infrastructure would provide a starting point, and would 
be sufficient to greatly improve researcher productivity on 
large, centrally managed datasets such as Genomics England.
Firstly, a \texttt{vcfzarrtools} command line utility which 
implements a subset of \texttt{bcftools} functionality (such as 
\texttt{view} and \texttt{query}) would immediately speed up 
limited region and field specific queries by orders of magnitude,
and would provide compatibility with existing workflows.
Developing such a utility would be a non-trivial but routine
engineering task.

@@

[Random fragment to include somewhere]
Row-wise storage of the genotype matrix makes accessing 
sample haplotypes (i.e. the columns),
required for many
applications~\cite[e.g.][]{durbin2014efficient,kelleher2019inferring} 
[TODO add phasing, imputation]
inefficient. The simple approach of storing the genotype matrix
in rectangular chunks allows efficient access both by row \emph{and}
column. Similarly, these rectangular chunks capture patterns
both of similarity between samples and linkage disequilibrium~\cite{mcvean2019linkage} 
between sites, such that conventional compression schemes 
perform very well~\ref{fig-data-storage}.

The Python data science ecosystem is increasingly popular, 
with packages such as NumPy~\citep{harris2020array},
SciPy~\citep{virtanen2020scipy},
Pandas~\citep{mckinney2010data},
Jupyter~\citep{kluyver2016jupyter}
enabling much of contemporary science.

BioNumpy~\citep{rand2022bionumpy},

% Although the Hadoop ecosystem has been hugely successful
% in many domains, it has had a limited impact in genomics.
% [List some things that have been tried for e.g variant calling
% , like ADAM~\citep{nothaft2015rethinking}].
% [Briefly mention two approaches using Hadoop clusters
% ~\citep{boufea2017managing,fan2020variant}, to not very
% exiting results].
% Distributed analytics for large-scale data is a mature
% techology, with many different approaches. Hadoop~\citep{white2012hadoop}
% is the original method, and remains the basis of the
% dominant ecosystem. HDFS. Spark [CITATION?] lets us
% do more complex analytics. Parquet [CITATION?]
% Distributed computation has been the reality in the analysis
% of genetic variation for many years,
% where data is most often provided as per-chromosome VCFs.
% This chunking provides a fairly straightforward way to parallelise
% computations, using HPC schedulers to split work per-file.
% When more fine-grained parallelism is required, users must
% provide exact genome coordinates of the genome regions
% as part of their submission scripts. Combining results
% is a manual operation.
% Workflow engines such as
% Snakemake~\cite{koster2012snakemake,molder2021sustainable}
% and [TODO mention WDL, Cromwell, etc?] make such analyses
% much easier, but there is still a great deal of room for
% mistakes.
% Per-chromosome VCFs are not practical for biobank scale
% projects, however, and variant call data tends to be split
% into many (still large) files.
% For example, the VCFs for 150K UKB WGS data~\cite{halldorsson2022sequences}
% are provided in 50kb chunks~\cite{browning2023statistical}, resulting in
% hundreds of files per chromosome.
% % This is horrible writing, but the point is important.
% While these files provide a natural
% unit for distributing work, the details of how they
% are split differ and essentially there is
% no interoperability among large-scale datasets because the VCFs are
% so cumbersome that a separate orchestration layer is required to
% access the data in parallel.

@@

\section{Choosing default compressor settings}
To inform the choice of compression settings across different fields 
in VCF data, we analysed their effect on compression ratio on 
data from the 1000 Genomes project.

>>>>

\section{Background}
Variant Call Format (VCF) is the standard format for interchanging genetic
variation data~\citep{danecek2011variant}, encoding information about 
DNA sequence polymorphisms among a set of samples with associated 
quality control metrics and metadata. 
Originally defined specifically as a text file,
it has been refined and standardised~\citep{rehm2021ga4gh} and the 
underlying data-model is now deeply embedded in bioinformatics practice.
Dataset sizes have grown explosively since the introduction of 
VCF as part of 1000 Genomes project~\citep{10002015global},
with Biobank scale initiatives such as 
Genomics England~\cite{turnbull2018100},
UK Biobank~\citep{bycroft2018genome,backman2021exome,halldorsson2022sequences,uk2023whole},
and the All of Us research program~\citep{all2024genomic}
collecting genome sequence data for hundreds of thousands of humans.
VCF's simple text-based design and widespread
support~\cite{garrison2022spectrum} makes it an 
excellent archival format, but it is an inefficient basis for analysis.
Methods that require efficient access to genotype data
either require conversion to the
PLINK~\cite{purcell2007plink,chang2015second} 
or BGEN~\citep{band2018bgen} 
formats~\citep[e.g.][]{yang2011gcta,mbatchou2021computationally,loh2015efficient}
or use 
bespoke binary formats~\citep[e.g.][]{
% Uses custom "bref3" format,
% https://faculty.washington.edu/browning/beagle/bref3.24May18.pdf
browning2018one, 
% .samples Zarr format
kelleher2019inferring,
% Has a "xcftools" package, but it still looks pretty experimental
hofmeister2023accurate} 
that support the required access patterns.
While PLINK and BGEN support efficient access to genotype data, neither
can accommodate the full flexibility of the VCF data model and conversion
is lossy.
PLINK's approach of storing the genotype matrix in uncompressed
packed-binary format provides efficient access to genotype 
data, but file sizes are substantially larger than 
the equivalent compressed VCF. 
For example, at two bits 
per genotype, the full genotype matrix for the GraphTyper SNP dataset 
in the 500K UKB WGS data~\citep{uk2023whole} is 116 TiB.
% 1,037,556,156 SNPs x  490,640 samples
% humanize.naturalsize(1_037_556_156 * 490_640 / 4, binary=True)
% '115.7 TiB'

Processing of Biobank scale datasets can be split into a 
few broad categories. The most basic analysis 
is quality control (QC). Variant QC is an 
involved and multi-faceted 
task [CITATIONS?], often requiring interactive, exploratory analysis
and incurring substantial computation over multiple QC fields.
Genotype calls are often refined via statistical methods,
% methods~\citep[e.g.][MORE and CHECK]{browning2021fast,browning2023statistical}
for example by phasing~[CITES] and imputation[CITES], 
creating additional dataset copies.
A common task to perform after QC is 
a genome wide association study (GWAS)~\cite{uffelmann2021genome}. 
The majority of tools for 
performing GWAS and related analyses require
data to be in PLINK format~\cite[e.g][]{chang2015second} [MORE],
and so data must be ``hard-called'' according to some QC criteria
and exported to additional copies.
Finally, variation datasets are often queried in exploratory
analyses, to find regions or samples of interest for a particular 
study~\cite[e.g.][]{chen2024novo}.

VCF cannot support any of these workflows efficiently
at the Biobank scale, and workarounds of copying subsets 
of data (i.e., discarding QC metrics in VCF) or copying to 
PLINK format have reached their limits.
The most intrinsically limiting aspect of VCF's design 
is its row-wise layout of data, which means that (for example)
information for a particular sample or field cannot be obtained without
retrieving the entire dataset.
The file-oriented paradigm is also unsuited to the realities 
of modern datasets, which are too large to download and 
often required to stay in-situ by data-access agreements.
Large files are currently stored in cloud environments, where the 
file systems that are required by classical file-oriented tools
are expensively emulated on the basic building blocks
of object storage. 
These multiple layers of inefficiences around processing
VCF data at scale in the cloud mean that it is 
time-consuming and expensive, and these vast datasets are 
not utilised to their full potential.

To achieve this full potential we 
need a new generation of tools that operate directly
on a primary data representation that supports 
efficient access across a range of applications,
with native support for cloud object storage.
Such a representation can be termed ``analysis-ready''
and ``cloud-native''~\citep{abernathey2021cloud}.
For the representation to be FAIR~\citep{wilkinson2016fair},
it must also be \emph{accessible}, using protocols that are 
``open, free, and universally implementable''.
There is currently no efficient, FAIR representation of genetic variation
data suitable for cloud deployments.
Hail~\cite{ganna2016ultra,hail2024} has become the dominant platform
for quality control of large-scale variation datasets, 
and has been instrumental in the delivery of datasets such as 
gnomadAD~\cite{karczewski2020mutational,chen2024genomic}.
While Hail is built on open components
from the Hadoop distributed computing ecosytem~\citep{white2012hadoop},
it is complex, vertically integrated software
with custom storage formats that are not intended for reuse.
Similarly, commerical solutions that have emerged to facilitate
the analysis of large-scale genetic variation data are either
based on proprietary~[Terra? BigQuery?]
or single-vendor technologies~\cite[e.g.][]{tiledb2024}.
The next generation of VCF analysis methods requires
an open, free and transparent data representation 
with multiple independent implementations.

In this article, we decouple the VCF data model from its row-oriented
file definition, and show how the data can be 
compactly stored and efficiently analysed in a cloud-native, FAIR manner.
We do this by translating VCF data into Zarr format,
a method of storing large-scale multidimensional data as a regular
grid of compressed chunks. 
Zarr's elegant simplicity and first-class support for 
cloud object stores have led to 
it gaining substantial traction
across the sciences, and it is now used in multiple petabyte-scale
datasets in cloud deployments (see Methods for details).
We present the VCF Zarr specification that formalises this 
mapping, and the \texttt{vcf2zarr} 
utility to reliably convert large-scale VCFs to Zarr.
We show that VCF Zarr is much more compact than 
VCF and is competitive with state-of-the-art
file-based VCF compression tools. 
Moreover, we show that Zarr's storage of data in an analysis-ready 
format greatly facilitates computation,
with various benchmarks being substantially faster than
\texttt{bcftools} based pipelines, and again competetive
with state-of-the-art file-oriented methods. Finally, we show the 
utility of VCF Zarr on the Genomics England aggV2 dataset,
demonstrating that common \texttt{bcftools} queries can be performed orders
of magnitude more quickly using simple Python scripts.

@@

\section{Discussion}
% The introduction of VCF as part of the work of the 1000 Genomes
% project~\cite{10002015global} has transformed a fragmented 
% landscape of mutually incompatible per-program file formats
% into an ecosystem of tools that largely
% interoperate~\cite{garrison2022spectrum}. In this 
% sense VCF has been a huge success, and the underlying data model
% is now deeply embedded in bioinformatics practice. As we have 
% illustrated here, the shortcomings of the VCF and BCF file
% formats (and more generally \emph{any} row-wise encoding of the 
% data model) in terms of computational accessibility are severe,
% and there can be little doubt that these file formats 
% are incurring a large economic cost and 
% holding back progress. 
% As we have illustrated in the Genomics England
% example here, these shortcomings are most acutely felt in 
% large-scale data human datasets but any setting in which we 
% wish to extract columns of the data from a row-wise 
% VCF format will be inefficient 
% when datasets are larger than a few hundred megabytes.

% As we have demonstrated, the columnar approach, based on the 
% popular Zarr standard and widely used compression libraries, 
% can support many different types of data access
% patterm efficiently, and has excellent computational accessibility.
% In this approach there are some significant departures from 
% standard practice, which warrant some discussion here.

% Perhaps the largest deviation in the approach that we advocate here 
% from standard practice is to decouple our concept of a dataset from
% a file. Each VCF is broken up into a set of Zarr arrays, each 
% corresponding to a large number of chunks stored in a directory
% hierarchy (Zarr also supports other forms of chunk storage 
% including single files and key-value stores however; see Methods).
% While there are some disadvantages to large numbers of small files,
% we argue that the benefits greatly outweigh them, particularly 
% in the large, centrally controlled dataset setting. 
% Firstly, this approach is inherently parallelisable, and chunks across the 
% dataset written and read in parallel without needing complex 
% synchronisation and prior layout planning. 
% Such parallel write access is very difficult 
% in the single-file setting, and one of the key weaknesses 
% of the HDF5 format~\citep{folk2011overview} [CITATION with 
% parallel troubles from HDF5?]
% Secondly, mapping chunks in arrays to individually addressed
% keys maps well to large-scale cloud object storage. Indeed,
% Zarr is specifically designed for such usage, and 
% several petabyte scale datasets have been successfully deployed
% and used [citations?].
% We have focused here on the classical single-server and 
% HPC cluster settings, but a major advantage of Zarr is 
% its cloud readiness.

% Probably getting into the weeds too much.
% Some large cloud-hosted VCF datasets are currently stored in 
% filesystems that emulate POSIX semantics on top of such object 
% stores, at a substantial cost in performance. Directly hosting 
% these datasets in object stores would have substantial 
% benefits, both in terms of storage costs and performance.
% Thirdly, breaking up fields into different directories (or 
% cloud buckets) enables fine-grained access control to the 
% dataset. For example, all users may be allowed access to
% variant-level data, but only a subset permitted to access
% call-level individual data. Such access controls can be 
% straightforwardly applied using file or cloud-hosting 
% methods on the relevant directores/buckets. Such concerns
% are entirely decoupled from the file format.

% This is poorly explained
Along with the move away from the single-file-storage paradigm,
the approach we outline here also suggests a shift from 
the classical Unix pipeline-based processing approach 
(but see below for a discussion of how existing pipelines can
be supported). Datasets flowing through pipelines inherently 
implies creating muliple copies of the data, and when 
multiple terabytes of either unrelated or unchanged data 
get copied several times through these pipes large inefficiencies 
are incurred (see Fig~\ref{fig-whole-matrix-compute}). 
We may think of this strategy as ``copy-oriented'' workflows.
A more efficient way to structure such multilayered computations
is in a  ``mask-oriented'' way. For example, to implement some
complex multilayered call filtering strategy we could imagine sequentially
updating a given per-call mask. Masks are cheap (e.g., the 
standard genotype missingness mask in Table~\ref{tab-genomics-england-data}
uses less than 1\% of the overall storage).

@@

Nonetheless, at least in the single-shared-copy model currently
used by most large-scale variation datasets, Zarr provides
an excellent basis for building a new generation of tools 
that compute \emph{directly} on the data, efficiently
accessing data by variant and by sample.
[Mention ``Cloud native data repositories''~\cite{abernathey2021cloud}]

The VCF Zarr specification and scalable vcf2zarr conversion utility
that we have provided here are the starting points of such 
an ecosystem, but much more would be needed to make a viable 
alternative for everyday bioinformatics workflows. However, a few key
pieces of infrastructure would provide a starting point, and would 
be sufficient to greatly improve researcher productivity on 
large, centrally managed datasets such as Genomics England.
Firstly, a \texttt{vcfzarrtools} command line utility which 
implements a subset of \texttt{bcftools} functionality (such as 
\texttt{view} and \texttt{query}) would immediately speed up 
limited region and field specific queries by orders of magnitude,
and would provide compatibility with existing workflows.
Developing such a utility would be a non-trivial but routine
engineering task.

@@

The Python data science ecosystem, 
built on the foundations of efficient array-oriented 
computing~\citep{harris2020array},
has a rich suite of powerful tools~\cite{
mckinney2010data,
rocklin2015dask,
kluyver2016jupyter,
hoyer2017xarray,
virtanen2020scipy} and gaining increasing
popularity in recent biological applications~\citep[e.g.][]{
abdennur2020cooler,
rand2022bionumpy,
open2c2024bioframe,
hou2024admix}.
Zarr has the potential to act as a unified, cloud-native 
efficient storage layer, freeing the next generation of 
bioinformatics tools
from the traditional burdens of text parsing and format
conversion.

% Zarr has also seen recent uptake 
% in single-cell single-cell genomics~\citep{dhapola2022scarf,virshup2023scverse}
% and multimodal spatial omics
% data~\citep{marconato2024spatialdata,baker2023emobject}.
% Recent additions using Zarr include the application of deep learning
% models to genomic sequence data~\citep{klie2023predictive} 
% and a browser for genetic variation data~\citep{konig2023divbrowse}.

% Workflow engines such as
% Snakemake~\cite{koster2012snakemake,molder2021sustainable}
% and [TODO mention WDL, Cromwell, etc?] make such analyses
% much easier, but there is still a great deal of room for
% mistakes.
% Per-chromosome VCFs are not practical for biobank scale
% projects, however, and variant call data tends to be split
% into many (still large) files.
% For example, the VCFs for 150K UKB WGS data~\cite{halldorsson2022sequences}
% are provided in 50kb chunks~\cite{browning2023statistical}, resulting in
% hundreds of files per chromosome.
% % This is horrible writing, but the point is important.
% While these files provide a natural
% unit for distributing work, the details of how they
% are split differ and essentially there is
% no interoperability among large-scale datasets because the VCFs are
% so cumbersome that a separate orchestration layer is required to
% access the data in parallel.

@@

\subsection{Choosing default compressor settings}
To inform the choice of compression settings across different fields 
in VCF data, we analysed their effect on compression ratio on 
data from the 1000 Genomes project.
[TODO, reffing Fig~\ref{fig-compression-shuffle} and other fig
on chunk size.]
